---
title: "Automatic detection of niche in Visium HD data using SpatialNiche"
author: "Hasan Al Reza"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    theme: united
mainfont: Arial
vignette: >
  %\VignetteIndexEntry{Automatic detection of niche in Visium HD data using SpatialNiche}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>",
  root.dir = './'
)
#knitr::opts_chunk$set(eval = FALSE)
```


This vignette outlines the steps of inference, analysis and visualization of niche for **a Visium HD spatial dataset using SpatialNiche**. We showcase SpatialNiche’s diverse functionalities by applying it to a Visium HD data on cells from Lymphangioleiomyomatosis (LAM, diseased) human lung from a patient.  

SpatialNiche requires raw Visium HD spaceranger output and optionally processed gene expression data of cells/bins from Seurat objects as the user input and models the niche based on a specific cell type or gene expression. 

Upon infering the niche, SpatialNichet provides functionality for further data exploration, analysis, and visualization. 


## Load the required libraries
```{r message=FALSE,warning=FALSE}
library(SpatialNiche)
library(ggplot2) #for visualization
library(dplyr) #for data wrangling
library(scattermore) #for spatial plotting
```

# Part I: Data input & processing and initialization of SpatialNiche list containing barcodes and images tibble
SpatialNiche requires two user inputs: one is the path to outs directory of the spaceranger output of the Visium HD dataset, and the other is the bins size you want to load. 

## Starting from spaceranger output
```{r}
localdir <- "../../../4_em/outs" # path to outs directory of the spaceranger output of the Visium HD dataset

DataHD<-Load_Spatial_Data(localdir, size = '008um') #loading the data using the path and the bin size

bcsHD<-DataHD$barcodes #pulling the barcodes tibble out for easier access

head(bcsHD) #look at the top entries

```


# Part II: Adding dconvolution information to SpatialNiche barcodes tibble
SpatialNiche supports import of spacexr's RCTD based deconvolution information. Thre RCTD result can be directly loaded as an rds object and imported directly into the barcodes tibble.

## Starting from RCTD output
```{r}
DeconvolutionHD<-Load_RCTD('../../../RCTD_d4_SCT.rds') #load rds object of RCTD results

bcsHD<-Add_Deconvolution_Info(bcsHD,DeconvolutionHD,addWeights=FALSE) #use in-built function to import RCTD deconvolution results

bcsHD<-bcsHD %>% na.omit() #remove blank barcodes with no deconvolution information

head(bcsHD[, c(1,13:15)]) #look at the top entries with deconvolution information

```

# Part III: Creating niche information
SpatialNiche creates niche information based on a specific cell type or gene expression.The niche information is defined based on a sepcific distance away from the bins/cells of interest.

## Defining niche based on a cell type
```{r}
PeripheryBCs<-Get_Periphery(barcodes=bcsHD, celltype='LAMCORE-1',distance=50,data.dir = localdir) #find barcodes in the cell specific niche

bcsHD$Periphery<-NA #assign empty column to barcodes tibble

bcsHD$Periphery[bcsHD$barcode%in%PeripheryBCs]<-"50 micron" #assign name to bins in the niche 

bcsHD$Periphery[is.na(bcsHD$Periphery)]<-"Tissue" #assign name to bins outside the niche 

bcsHD$Periphery[bcsHD$Periphery=="Tissue" & bcsHD$DeconvolutionLabel1=='LAMCORE-1']<-'LAMCORE-1' #assign name to bins specifying the niche 

head(bcsHD[, c(1,14:16)]) #look at the top entries with cell based niche information

```

## Defining niche based on gene expression
```{r}
em_d4 <- readRDS("../../../em_d4_SCT.rds") #import seurat object

bcsHD<-Add_Expression(bcsHD,em_d4,c("PMEL")) #add gene expression data to barcodes tibble

PeripheryGCs<-Get_Gene_Periphery(barcodes=bcsHD, gene=c("PMEL"), seurat=em_d4,distance=50,data.dir = localdir) #find barcodes in the gene specific niche

bcsHD$Peripheryg<-NA #assign empty column to barcodes tibble

bcsHD$Peripheryg[bcsHD$barcode%in%PeripheryGCs]<-"50 micron" #assign name to bins in the niche 

bcsHD$Peripheryg[is.na(bcsHD$Peripheryg)]<-"Tissue" #assign name to bins outside the niche 

bcsHD$Peripheryg[bcsHD$Peripheryg=="Tissue" & bcsHD$PMEL>0]<-'PMEL' #assign name to bins specifying the niche 

head(bcsHD[, c(1,14:18)]) #look at the top entries with gene expression based niche information

```


# Part IV: Visualizing niche results
Upon infering the niche information, SpatialNiche provides various functionality for further data exploration, analysis, and visualization. Specifically:

* It provides utilities for slicing the dataset into smaller region of interest.

* It provides tools for visualization of deconvolution information.

* It provides tools for visualization of niche information.

## slicing the dataset into smaller ROIs
```{r}
SliceA<-Get_Square('s_008um_00128_00230-1',250,bcsHD) #find barcodes in a Square ROI

head(SliceA) #look at the top barcodes in the Square ROI

SliceB<-Get_Rectangle('s_008um_00331_00466-1',300,0.65,1.1,bcsHD) #find barcodes in a Rectangle ROI, use xfactor and yfactor to control size of the rectangle

head(SliceB) #look at the top barcodes in the Rectangle ROI

SliceC<-Get_Circle('s_008um_00128_00230-1',125,bcsHD, localdir) #find barcodes in a Circle ROI

head(SliceC) #look at the top barcodes in the Circle ROI

```

## Visualizing based on deconvolution information
```{r}
bcsHD %>% filter(DeconvolutionClass=="singlet") %>%
  ggplot(aes(x = imagecol_scaled, y = -imagerow_scaled,color=DeconvolutionLabel1)) +
  Geom_Spatial(data = DataHD$images, aes(grob = grob), x = 0.5, y = 0.5)+
  geom_scattermore(pointsize = 2,pixels = rep(2000,2))+
  coord_cartesian(expand = FALSE) +
  xlab("") +
  ylab("") +
  theme_set(theme_bw(base_size = 10))+
  theme_minimal() +
  theme(axis.text = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) #Visualize singlets by deconvolution label
bcsHD  %>%
  ggplot(aes(x = imagecol_scaled, y = -imagerow_scaled,color=DeconvolutionLabel1)) +
  Geom_Spatial(data = DataHD$images, aes(grob = grob), x = 0.5, y = 0.5) +
  geom_point(shape=16,size=2)+
  coord_cartesian(expand = FALSE) +
  xlab("") +
  ylab("") + 
  theme_set(theme_bw(base_size = 10))+
  theme_minimal() +
  theme(axis.text = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) #Visualize all bins by deconvolution label

```

## Visualizing based on niche information
```{r}
bcA=bcsHD[bcsHD$Periphery!='Tissue',] #Create new tibble with only bins in the cell specific niche

bcA %>%
  ggplot(aes(x = imagecol_scaled, y = -imagerow_scaled,color=Periphery)) +
  Geom_Spatial(data = DataHD$images, aes(grob = grob), x = 0.5, y = 0.5) +
  geom_scattermore(pointsize = 2,pixels = rep(2000,2)) +
  scale_color_manual(values=c('green','red','grey')) +
  coord_cartesian(expand = FALSE) +
  xlab("") +
  ylab("") +
  theme_set(theme_bw(base_size = 10)) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) #Visualize bins in the niche
bcA=bcsHD[bcsHD$Peripheryg!='Tissue',] #Create new tibble with only bins in the gene specific niche

bcA %>%
  ggplot(aes(x = imagecol_scaled, y = -imagerow_scaled,color=Peripheryg)) +
  Geom_Spatial(data = DataHD$images, aes(grob = grob), x = 0.5, y = 0.5) +
  geom_scattermore(pointsize = 2,pixels = rep(2000,2)) +
  scale_color_manual(values=c('green','red','grey')) +
  coord_cartesian(expand = FALSE) +
  xlab("") +
  ylab("") +
  theme_set(theme_bw(base_size = 10)) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank()) #Visualize bins in the niche

```


# Part V: Calculating distance information
SpatialNiche is also capable of creating distance information by translating the pixel sizes to the micron scale. The results are output as a distane matrix that can be converted to a dataframe for use.

## Defining niche based on a cell type
```{r}
em_d4G1 <- readRDS("../../../KWB/em_d4_G1.rds") #import seurat object

spatial.locs = Seurat::GetTissueCoordinates(em_d4G1, scale = NULL, cols = c("imagerow", "imagecol")) #get coordinates from seurat object

spatial.locs = spatial.locs[,1:2] #remove barcodes from coordinates table

spot.size = 8 #assigning spot size

conversion.factor = spot.size/em_d4G1@images$slice1.008um@scale.factors$spot #calculate conversion factor

spatial.factors = data.frame(ratio = conversion.factor, tol = spot.size/2) #creating spatial factor

d.spatial <- Compute_Cell_Distance(coordinates = spatial.locs, ratio = spatial.factors$ratio, tol = spatial.factors$tol) #creating distance matrix

min(d.spatial[d.spatial!=0]) # this value should approximately equal 8um for 10X VisiumHD data

df=as.data.frame(as.matrix(d.spatial)) #convert distance matrix to dataframe

head(df[,1:6]) #look at the distance between the top barcodes

```


# Part VI: Save the SpatialNiche files
```{r eval=FALSE}
saveRDS(DataHD, 'DataHD.rds') #save SpatialNiche list containing barcodes and images tibble

saveRDS(PeripheryBCs, 'PeripheryBCs.rds') #save barcodes in the cell specific niche

saveRDS(PeripheryGCs, 'PeripheryGCs.rds') #save barcodes in the gene specific niche
```


```{r}
sessionInfo()
```
