---
title: "Automatic detection of niche in Visium HD data using SpatialNiche"
author: "Hasan Al Reza"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    theme: united
mainfont: Arial
vignette: >
  %\VignetteIndexEntry{Automatic detection of niche in Visium HD data using SpatialNiche}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>",
  root.dir = './'
)
#knitr::opts_chunk$set(eval = FALSE)
```


This vignette outlines the steps of inference, analysis and visualization of niche for **a Visium HD spatial dataset using SpatialNiche**. We showcase SpatialNiche’s diverse functionalities by applying it to a Visium HD data on cells from Lymphangioleiomyomatosis (LAM, diseased) human lung from a patient.  

SpatialNiche requires raw Visium HD spaceranger output and optionally processed gene expression data of cells/bins from Seurat objects as the user input and models the niche based on a specific cell type or gene expression. 

Upon infering the niche, SpatialNichet provides functionality for further data exploration, analysis, and visualization. 


## Load the required libraries
```{r message=FALSE,warning=FALSE}
library(SpatialNiche)
library(ggplot2)
library(dplyr)
library(scattermore)
```

# Part I: Data input & processing and initialization of SpatialNiche list containing barcodes and images tibble
SpatialNiche requires two user inputs: one is the path to outs directory of the spaceranger output of the Visium HD dataset, and the other is the bins size you want to load. 

## Starting from spaceranger output
```{r}
ptm = Sys.time()

localdir <- "../../../4_em/outs" # path to outs directory of the spaceranger output of the Visium HD dataset

DataHD<-Load_Spatial_Data(localdir, size = '008um') #loading the data using the path and the bin size

bcsHD<-DataHD$barcodes #pulling the barcodes tibble out for easier access

head(bcsHD) #look at the top entries

```


# Part II: Adding dconvolution information to SpatialNiche barcodes tibble
SpatialNiche supports import of spacexr's RCTD based deconvolution information. Thre RCTD result can be directly loaded as an rds object and imported directly into the barcodes tibble.

## Starting from RCTD output
```{r}
ptm = Sys.time()

DeconvolutionHD<-Load_RCTD('../../../RCTD_d4_SCT.rds') #load rds object of RCTD results

bcsHD<-Add_Deconvolution_Info(bcsHD,DeconvolutionHD,addWeights=FALSE) #use in-built function to import RCTD deconvolution results

bcsHD<-bcsHD %>% na.omit() # remove blank barcodes with no deconvolution information

head(bcsHD[, c(1,13:15)]) #look at the top entries with deconvolution information

```

# Part III: Creating niche information
SpatialNiche creates niche information based on a specific cell type or gene expression.The niche information is defined based on a sepcific distance away from the bins/cells of interest.

## Defining niche based on a cell type
```{r}
ptm = Sys.time()

PeripheryBCs<-Get_Periphery(barcodes=bcsHD, celltype='LAMCORE-1',distance=50,data.dir = localdir)

bcsHD$Periphery<-NA
bcsHD$Periphery[bcsHD$barcode%in%PeripheryBCs]<-"50 micron"
bcsHD$Periphery[is.na(bcsHD$Periphery)]<-"Tissue"
bcsHD$Periphery[bcsHD$Periphery=="Tissue" & bcsHD$DeconvolutionLabel1=='LAMCORE-1']<-'LAMCORE-1'

head(bcsHD[, c(1,14:16)]) #look at the top entries with cell based niche information

```

## Defining niche based on gene expression
```{r}
ptm = Sys.time()

em_d4 <- readRDS("../../../em_d4_SCT.rds")

bcsHD<-Add_Expression(bcsHD,em_d4,c("PMEL"))

PeripheryGCs<-Get_Gene_Periphery(barcodes=bcsHD, gene=c("PMEL"), seurat=em_d4,distance=50,data.dir = localdir)

bcsHD$Peripheryg<-NA
bcsHD$Peripheryg[bcsHD$barcode%in%PeripheryGCs]<-"50 micron"
bcsHD$Peripheryg[is.na(bcsHD$Peripheryg)]<-"Tissue"
bcsHD$Peripheryg[bcsHD$Peripheryg=="Tissue" & bcsHD$PMEL>0]<-'PMEL'

head(bcsHD[, c(1,14:18)]) #look at the top entries with gene expression based niche information

```


# Part IV: Visualizing niche results
Upon infering the niche information, SpatialNiche provides various functionality for further data exploration, analysis, and visualization. Specifically:

* It provides utilities for slicing the dataset into smaller region of interest.

* It provides tools for visualization of deconvolution information.

* It provides tools for visualization of niche information.

## slicing the dataset into smaller ROIs
```{r}
SliceA<-Get_Square('s_008um_00128_00230-1',250,bcsHD)

head(SliceA)

SliceB<-Get_Rectangle('s_008um_00331_00466-1',300,0.65,1.1,bcsHD)

head(SliceB)

SliceC<-Get_Circle('s_008um_00128_00230-1',125,bcsHD, localdir)

head(SliceC)

```

## Visualizing based on deconvolution information
```{r}
bcsHD %>% filter(DeconvolutionClass=="singlet") %>%
  ggplot(aes(x = imagecol_scaled, y = -imagerow_scaled,color=DeconvolutionLabel1)) +
  Geom_Spatial(data = DataHD$images, aes(grob = grob), x = 0.5, y = 0.5)+
  geom_scattermore(pointsize = 2,pixels = rep(2000,2))+
  coord_cartesian(expand = FALSE) +
  xlab("") +
  ylab("") +
  theme_set(theme_bw(base_size = 10))+
  theme_minimal() +
  theme(axis.text = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())+
  #geom_rect(aes(xmin=DFRectA$Xmin, xmax=DFRectA$Xmax, ymin=DFRectA$Ymax, ymax=DFRectA$Ymin),fill=NA,color="black",linewidth=1)+
  #geom_circle(aes(x0=DFRectC$Xavg, y0=DFRectC$Yavg, r=DFRectC$rad),fill=NA,color="black",linewidth=1)+
  #scale_color_manual(values=ColorPalette())
  NoLegend()

bcsHD  %>%
  ggplot(aes(x = imagecol_scaled, y = -imagerow_scaled,color=DeconvolutionLabel1)) +
  Geom_Spatial(data = DataHD$images, aes(grob = grob), x = 0.5, y = 0.5) +
  geom_point(shape=16,size=2)+ #sahpe 15=square 16=circle
  coord_cartesian(expand = FALSE) +
  #scale_color_manual(values = ColorPalette(),na.value = NA) +
  xlab("") +
  ylab("") + theme_set(theme_bw(base_size = 10))+
  theme_minimal() +
  theme(axis.text = element_blank())#+NoLegend()

```

## Visualizing based on niche information
```{r}
bcA=bcsHD[bcsHD$Periphery!='Tissue',]

bcA %>%
  ggplot(aes(x = imagecol_scaled, y = -imagerow_scaled,color=Periphery)) +
  Geom_Spatial(data = DataHD$images, aes(grob = grob), x = 0.5, y = 0.5) +
  geom_scattermore(pointsize = 2,pixels = rep(2000,2)) +
  scale_color_manual(values=c('green','red','grey')) +
  coord_cartesian(expand = FALSE) +
  xlab("") +
  ylab("") +
  theme_set(theme_bw(base_size = 10)) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())

bcA=bcsHD[bcsHD$Peripheryg!='Tissue',]

bcA %>%
  ggplot(aes(x = imagecol_scaled, y = -imagerow_scaled,color=Peripheryg)) +
  Geom_Spatial(data = DataHD$images, aes(grob = grob), x = 0.5, y = 0.5) +
  geom_scattermore(pointsize = 2,pixels = rep(2000,2)) +
  scale_color_manual(values=c('green','red','grey')) +
  coord_cartesian(expand = FALSE) +
  xlab("") +
  ylab("") +
  theme_set(theme_bw(base_size = 10)) +
  theme_minimal() +
  theme(axis.text = element_blank(),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())

```


# Part V: Calculating distance information
SpatialNiche is also capable of creating distance information by translating the pixel sizes to the micron scale. The results are output as a distane matrix that can be converted to a dataframe for use.

## Defining niche based on a cell type
```{r}
ptm = Sys.time()

em_d4G1 <- readRDS("../../../KWB/em_d4_G1.rds")

spatial.locs = Seurat::GetTissueCoordinates(em_d4G1, scale = NULL, cols = c("imagerow", "imagecol"))
spatial.locs = spatial.locs[,1:2]

spot.size = 8

conversion.factor = spot.size/em_d4G1@images$slice1.008um@scale.factors$spot
spatial.factors = data.frame(ratio = conversion.factor, tol = spot.size/2)

d.spatial <- Compute_Cell_Distance(coordinates = spatial.locs, ratio = spatial.factors$ratio, tol = spatial.factors$tol)
min(d.spatial[d.spatial!=0]) # this value should approximately equal 8um for 10X VisiumHD data
df=as.data.frame(as.matrix(d.spatial))

head(df)
```


# Part VI: Save the SpatialNiche files
```{r eval=FALSE}
saveRDS(bcsHD, 'bcsHD.rds')
saveRDS(PeripheryBCs, 'PeripheryBCs.rds')
saveRDS(PeripheryGCs, 'PeripheryGCs.rds')
```


```{r}
sessionInfo()
```
